The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-05-30 Manuscript/dark.tex
Added reminder to cite other similar papers
Reworking introduction

Phylogenetic comparative methods (PCMs) were initially developed in the 1980s to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). PCMs have become extremely popular over recent years; in 2013 alone Harvey and Pagel's \citeyearpar{harvey1991comparative} book on \textit{The Comparative Method in Evolutionary Biology} was cited 195 times (Google Scholar, 22nd May 2014), and at the 2014 Evolution meeting in Raleigh NC, we estimate that over XX %need to count these!
talks used PCMs of one form or another.

PCMs also have a ``dark side''; they make various assumptions and suffer from biases in the same way as any statistical method. Unfortunately, the more popular these methods become, the less awareness methods-users seem to have of this ``dark side''. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording

>>>>

Phylogenetic comparative methods (PCMs) were initially developed a long time ago in a galaxy far, far away - also know as the 1980s - to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). PCMs have become extremely popular over recent years; in 2013 alone Harvey and Pagel's \citeyearpar{harvey1991comparative} book on \textit{The Comparative Method in Evolutionary Biology} was cited 195 times (Google Scholar, 22nd May 2014), and at the 2014 Evolution meeting in Raleigh NC, we estimate that over XX %need to count these!
talks used PCMs of one form or another. With advances in blah balh, there has never been a better time to be a comparative biologist.

However, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in the same way as any statistical method \citet{freckleton2009seven} Unfortunately, the more popular these methods become, the less awareness methods-users seem to have of this ``dark side''. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. Additionally, little consideration is given to whether using a PCM is really appropriate for the question \citet{losos2011seeing}. Although methods-developers (and advanced methods-users) are aware of these issues, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording

%some comments showing we're not the only people to think this, nor the first to say it.
%cite Boettiger etal 2012
%cite Losos 2011
%cite Freckleton 2009
%cite Blomberg et al 2012
